Using Path-Dependent Types to Build Type-Safe JavaScript Foreign Function Interfaces

### 2014-02-04 papers/icwe14/icwe14.tex
++i

\begin{abstract}
Encoding dynamically typed APIs in statically typed languages can be challenging. Incidentally, existing statically typed languages compiling to JavaScript expose its native API in ways that either are not type safe or give less expression power to developers. We present two new ways to encode the challenging parts of the Web browser API such that both type safety and expression power are preserved. Our first encoding relies on type parameters and can be implemented in most mainstream languages but drags phantom types up to the usage sites. The second encoding does not suffer from this inconvenience but requires dependent types.
\end{abstract}

@@

We recently observed the emergence of several statically typed programming languages compiling to JavaScript (\emph{e.g.} Java/GWT~\cite{Kereki09_GWT}, Dart~\cite{Griffith11_Dart}, TypeScript~\cite{fenton2012typescript}, Kotlin\footnote{\href{http://kotlin.jetbrains.org}{http://kotlin.jetbrains.org}}, Opa\footnote{\href{http://opalang.org}{http://opalang.org}}, SharpKit\footnote{\href{http://sharpkit.net}{http://sharpkit.net}}, Haxe~\cite{Cannasse08_HaXe}, Scala~\cite{Doeraene13_ScalaJs}, Idris~\cite{Brady13_Idris}, Elm~\cite{czaplicki2012elm}). Developers can use them to write the client-side part of their Web applications, relying on a \emph{foreign function interface} mechanism to call the Web browser native API.

However we observe that, despite these languages are statically typed, their integration of the browser API either is not type safe or gives less expression power. Indeed, integrating an API designed for a dynamically typed language into a statically typed language can be challenging. For instance, the \jscode{createElement} function return type depends on the value of its parameter: \jscode{createElement('div')} returns a \jscode{DivElement}, \jscode{createElement('input')} returns an \jscode{InputElement}, \emph{etc}.

@@

The remainder of the paper is organized as follows. The next section reviews the most common functions of the browser API and how they are integrated in statically typed languages. Section \ref{sec-contribution} shows ways to improve their integration. Section \ref{sec-validation} validates our contribution. Section \ref{sec-related} discusses some related works and section \ref{sec-conclusion} concludes.

@@

Listing \ref{lst-problem-cast} shows a Scala implementation of the \jscode{slideshow} program using an API following the standard specifications of HTML and DOM. The listing contains two type casts, making it possible to use the \scalacode{src} property on the \scalacode{img} value and the \scalacode{button} property on the \scalacode{event} value, respectively.

@@

We first propose a solution in the Java mainstream language, using \emph{generics}, and show how we can improve it using \emph{path-dependent types}, in Scala.

@@

\begin{enumerate}
 \item Define a parametrized class \scalacode{P<U>}
 \item Set the type of \scalacode{p} to \scalacode{P<U>}
 \item Use type \scalacode{U} instead of type \scalacode{T}
 \item Define as many values of type \scalacode{P<U>} as there are possible values for \scalacode{p}, each one fixing its \scalacode{U} type parameter to the corresponding more precise type
\end{enumerate}

Listing~\ref{lst-generics-dom} shows this approach applied to the \jscode{createElement} function which return type depends on its \jscode{name} parameter value: a type \scalacode{ElementName<E>} has been created, the type of the \scalacode{name} parameter has been set to \scalacode{ElementName<E>} instead of \scalacode{String}, and the return type of the function is \scalacode{E} instead of \scalacode{Element}. The \scalacode{ElementName<E>} type encodes the relationship between the name of an element and the type of this element\footnote{The type parameter \scalacode{E} is also called a \emph{phantom type}~\cite{leijen1999domain} because \scalacode{ElementName} values never hold a \scalacode{E} value}. For instance, we created a value \scalacode{Input} of type \scalacode{ElementName<InputElement>}.

\begin{figure}
\begin{lstlisting}[label=lst-generics-dom,language=java,caption={Encoding of the \jscode{createElement} function using type parameters}]
class ElementName<E> {}

<E> E createElement(ElementName<E> name);

final ElementName<InputElement> Input = new ElementName<>();
final ElementName<ImageElement> Img = new ElementName<>();
// etc. for each possible element name
\end{lstlisting}
\end{figure}

\scalacode{getElementsByTagName} can be straightforwardly encoded in a similar way, and listing~\ref{lst-generics-events} shows the encoding of the \jscode{addEventListener} function. Our listing assumes the definition of a \scalacode{Function<A, B>} type representing a function from \scalacode{A} to \scalacode{B}.

\begin{figure}
\begin{lstlisting}[label=lst-generics-events,language=java,caption={Encoding of the \jscode{addEventListener} function using type parameters}]
class EventName<E> {}

interface EventTarget {
  <E> void addEventListener(
          EventName<E> name, Function<E, Void> callback);
}

final EventName<MouseEvent> Click = new EventName<>();
final EventName<KeyboardEvent> KeyUp = new EventName<>();
// etc. for each possible event name
\end{lstlisting}
\end{figure}

@@

Listing \ref{lst-slideshow-generics} illustrate the usage of such an encoding by implementing our \texttt{slideshow} program in Scala. Passing the \scalacode{Img} value as a parameter to the \scalacode{createElement} function fixes its \scalacode{E} type parameter to \scalacode{ImageElement} so the returned value has the most possible precise type and the \scalacode{src} property can be used on it. Similarly, passing the \scalacode{Click} value to the \scalacode{addEventListener} function fixes its \scalacode{E} type parameter to \scalacode{MouseEvent}, so the event listener has the most possible precise type and the \scalacode{button} property can be used on the \scalacode{event} parameter.

\begin{figure}
\begin{lstlisting}[label=lst-generics-comb,language=java,caption={Combination of \scalacode{getElementsByTagName} and \scalacode{addEventListener} functions encoded using type parameters}]
<A, B> void findAndListenTo(
    ElementName<A> tagName,
    EventName<B> eventName,
    Function2<A, B, Void> listener) {
  for (A element : document.getElementsByTagName(tagName)) {
    element.addEventListener(
         eventName,
         event -> listener.apply(event, element));
  }
}
\end{lstlisting}
\end{figure}

\begin{figure}
\begin{lstlisting}[label=lst-generics-react,language=java,caption={Partial application of \scalacode{addEventListener} encoded with type parameters}]
<A> Function<Function<A, Void>, Void> observe(
      EventTarget target, EventName<A> name) {
  return listener -> {
    target.addEventListener(name, listener);
  }
}
\end{lstlisting}
\end{figure}

Finally, listing \ref{lst-generics-comb} and \ref{lst-generics-react} show how the challenging functions of section \ref{sec-limitations}, \jscode{findAndListenTo} and \jscode{observe}, can be implemented with our encoding. They are basically a direct translation from JavaScript to Java.

Our encoding is type safe and gives as much expression power as the native API since it makes it possible to implement exactly the same functions as we are able to implement in plain JavaScript.

@@

This section shows how we can remove the extra type parameters needed in the previous section by using \emph{path-dependent types}. Essentially, the idea is that, instead of using a type parameter, we use a \emph{type member}. Listings \ref{lst-dt-dom} and \ref{lst-dt-events} show this encoding in Scala for \jscode{createElement} and \jscode{addEventListener}, respectively.

@@

def createElement(name: ElementName): name.Element

@@

Now, the \scalacode{ElementName} type has no type parameter but a type member \scalacode{Element}. The return type of the \scalacode{createElement} function is \scalacode{name.Element}: it refers to the \scalacode{Element} type member of its \scalacode{name} parameter. The \scalacode{Div} and \scalacode{Input} values illustrate how their corresponding element type is fixed.

@@

trait EventTarget {
  def addEventListener(name: EventName)
                      (handler: name.Event => Unit): Unit
}
\end{lstlisting}
\end{figure}

[FIXME un peu plus de details sur le fonctionnement des path-dependent types]

@@

@@

>>>>

\begin{abstract}
Statically typed programming languages compiling to JavaScript make it worth considering the benefits of static typing to write Web applications. Nevertheless, at some point these languages need to expose the Web browser dynamically typed native API, which is a challenging task. Indeed, we observe that existing statically typed languages compiling to JavaScript expose the browser API in ways that either are not type safe or give less expression power and therefore force developers to pick only one of these two properties. This article presents new ways to encode the challenging parts of the Web browser API in static type systems such that both type safety and expression power are preserved. Our first encoding relies on type parameters and can be implemented in most mainstream languages but drags phantom types up to the usage sites. The second encoding does not suffer from this inconvenience but requires dependent types.
\end{abstract}

@@

We recently observed the emergence of several statically typed programming languages compiling to JavaScript (\emph{e.g.} Java/GWT~\cite{Kereki09_GWT}, Dart~\cite{Griffith11_Dart}, TypeScript~\cite{fenton2012typescript}, Kotlin\footnote{\href{http://kotlin.jetbrains.org}{http://kotlin.jetbrains.org}}, Opa\footnote{\href{http://opalang.org}{http://opalang.org}}, SharpKit\footnote{\href{http://sharpkit.net}{http://sharpkit.net}}, Haxe~\cite{Cannasse08_HaXe}, Scala~\cite{Doeraene13_ScalaJs}, Idris~\cite{Brady13_Idris}, Elm~\cite{czaplicki2012elm}). These languages make it possible to benefit from static typing (\emph{e.g.} better refactoring support in IDE, earlier error detection) to write Web applications, usually written in the dynamically typed language JavaScript. Nevertheless, at some point developers need a way to interface with the underlying Web browser dynamically typed native API using a \emph{foreign function interface} mechanism.

We observe that, despite these languages are statically typed, their integration of the browser API either is not type safe or gives less expression power. Indeed, integrating an API designed for a dynamically typed language into a statically typed language can be challenging. For instance, the \jscode{createElement} function return type depends on the value of its parameter: \jscode{createElement('div')} returns a \jscode{DivElement}, \jscode{createElement('input')} returns an \jscode{InputElement}, \emph{etc}.

@@

The remainder of the paper is organized as follows. The next section reviews the most common functions of the browser API and how they are integrated in statically typed languages. Section \ref{sec-contribution} shows two ways to improve their integration such that type safety and expressivity are preserved. Section \ref{sec-validation} validates our contribution. Section \ref{sec-related} discusses some related works and section \ref{sec-conclusion} concludes.

@@

Listing \ref{lst-problem-cast} shows a Scala implementation of the \jscode{slideshow} program using an API following the standard specifications of HTML and DOM. The listing contains two type casts, needed to use the \scalacode{src} property on the \scalacode{img} value and the \scalacode{button} property on the \scalacode{event} value, respectively.

@@

Our listings use the Scala language, though our first solution could be implemented in any language with basic type parameters support, such as Java's \emph{generics}. Our second solution is an improvement of the first one, using \emph{path-dependent types}.

@@

\begin{enumerate}
 \item Define a parametrized class \scalacode{P[U]}
 \item Set the type of \scalacode{p} to \scalacode{P[U]}
 \item Use type \scalacode{U} instead of type \scalacode{T}
 \item Define as many values of type \scalacode{P[U]} as there are possible values for \scalacode{p}, each one fixing its \scalacode{U} type parameter to the corresponding more precise type
\end{enumerate}

\begin{figure}
\begin{lstlisting}[label=lst-generics-dom,language=Scala,caption={Encoding of the \jscode{createElement} function using type parameters}]
class ElementName[E]

trait Document {
  def createElement[E](name: ElementName[E]): E
  def getElementsByTagName[E](name: ElementName[E]): Array[E]
}

val Input = new ElementName[InputElement]
val Img = new ElementName[ImageElement]
// etc. for each possible element name
\end{lstlisting}
\end{figure}

Listing~\ref{lst-generics-dom} shows this approach applied to the \jscode{createElement} and \jscode{getElementsByTagName} functions which return type depends on their \jscode{name} parameter value: a type \scalacode{ElementName[E]} has been created, the type of the \scalacode{name} parameter has been set to \scalacode{ElementName[E]} instead of \scalacode{String}, and the return type of the function is \scalacode{E} instead of \scalacode{Element} (\scalacode{Array[E]} instead of \scalacode{Array[Element]}, in the case of \scalacode{getElementsByTagName}). The \scalacode{ElementName[E]} type encodes the relationship between the name of an element and the type of this element\footnote{The type parameter \scalacode{E} is also called a \emph{phantom type}~\cite{leijen1999domain} because \scalacode{ElementName} values never hold a \scalacode{E} value}. For instance, we created a value \scalacode{Input} of type \scalacode{ElementName[InputElement]}.

\begin{figure}
\begin{lstlisting}[label=lst-generics-events,language=Scala,caption={Encoding of the \jscode{addEventListener} function using type parameters}]
class EventName[E]

trait EventTarget {
  def addEventListener[E](
          name: EventName[E], callback: E => Unit): Unit
}

val Click = new EventName[MouseEvent]
val KeyUp = new EventName[KeyboardEvent]
// etc. for each possible event name
\end{lstlisting}
\end{figure}

Listing~\ref{lst-generics-events} shows the encoding of the \jscode{addEventListener} function. The \scalacode{EventName[E]} type represents the name of an event which type is \scalacode{E}. For instance, \scalacode{Click} is a value of type \scalacode{EventName[MouseEvent]}: when a user adds an event listener to the \scalacode{Click} event, it fixes to \scalacode{MouseEvent} the type parameter \scalacode{E} of the \scalacode{callback} function passed to \scalacode{addEventListener}.

@@

Listing \ref{lst-slideshow-generics} illustrates the usage of such an encoding by implementing our \texttt{slideshow} program in Scala. Passing the \scalacode{Img} value as a parameter to the \scalacode{createElement} function fixes its \scalacode{E} type parameter to \scalacode{ImageElement} so the returned value has the most possible precise type and the \scalacode{src} property can be used on it. Similarly, passing the \scalacode{Click} value to the \scalacode{addEventListener} function fixes its \scalacode{E} type parameter to \scalacode{MouseEvent}, so the event listener has the most possible precise type and the \scalacode{button} property can be used on the \scalacode{event} parameter. It is worth noting that there is no type cast in the listing because the browser API is exposed in a way that preserves enough type information.

\begin{figure}
\begin{lstlisting}[label=lst-generics-comb,language=Scala,caption={Combination of \scalacode{getElementsByTagName} and \scalacode{addEventListener} functions encoded using type parameters}]
def findAndListenTo[A, B](
      tagName: ElementName[A],
      eventName: EventName[B],
      listener: (A, B) => Unit) = {
  for (element <- document.getElementsByTagName(tagName)) {
    element.addEventListener(eventName, event => {
      listener(event, element)
    })
  }
}
\end{lstlisting}
\end{figure}

\begin{figure}
\begin{lstlisting}[label=lst-generics-react,language=Scala,caption={Partial application of \scalacode{addEventListener} encoded with type parameters}]
def observe[A](target: EventTarget, name: EventName[A]) = {
  (listener: A => Unit) => {
    target.addEventListener(name, listener)
  }
}
\end{lstlisting}
\end{figure}

Finally, listing \ref{lst-generics-comb} and \ref{lst-generics-react} show how the challenging functions of section \ref{sec-limitations}, \jscode{findAndListenTo} and \jscode{observe}, can be implemented with our encoding. They are basically a direct translation from JavaScript syntax to Scala syntax, with additional type annotations. Our encoding is type safe and gives as much expression power as the native API since it is possible to implement exactly the same functions as we are able to implement in plain JavaScript.

@@

This section shows how we can remove the extra type parameters needed in the previous section by using \emph{path-dependent types}~\cite{Odersky03_vObj}. Essentially, the idea is to model type parameters using \emph{type members}, as suggested in~\cite{odersky2005scalable}.

@@

trait Document {
  def createElement(name: ElementName): name.Element
  def getElementsByTagName(
        name: ElementName): Array[name.Element]
}

@@

@@

trait EventTarget {
  def addEventListener(name: EventName)
                      (callback: name.Event => Unit): Unit
}
\end{lstlisting}
\end{figure}

Listings \ref{lst-dt-dom} and \ref{lst-dt-events} show this encoding in Scala for \jscode{createElement}, \jscode{getElementsByTagName} and \jscode{addEventListener}. Now, the \scalacode{ElementName} type has no type parameter but a type member \scalacode{Element}. The return type of the \scalacode{createElement} function is \scalacode{name.Element}: it refers to the \scalacode{Element} type member of its \scalacode{name} parameter. The \scalacode{Div} and \scalacode{Input} values illustrate how their corresponding element type is fixed: if one writes \scalacode{createElement(Input)}, the return type is the \scalacode{Element} type member of the \scalacode{Input} value, namely \scalacode{InputElement}. The same idea applies to \scalacode{EventName} and \scalacode{addEventListener}: the name of the event fixes the type of the callback.

@@

Mesurer les gains.

@@

Christian Hofer \emph{et. al.} leveraged the fact that type parameters can be modeled with type members~\cite{hofer2008polymorphic}.
